---
title: "R Notebook"
output: html_notebook
---
---
title: "R Notebook"
output: html_notebook
---
 
This workbook is for plotting of cover data for Norfolk Island Ecosystem Health Surveys

Read in data

# NOTE - script will need to be updated reflecting new labels 



```{r}
library(tidyverse)
library(ggplot2)

master_data <- read.csv("CORAL_NET_MASTER_April_23_UPDATED.csv")
```

Transpose data frame

```{r}

# Transpose the data frame
t.cover <- master_data %>% gather(Genera, Cover, Acanthastrea.Bleached:Black_turf)
str(t.cover)

```

Re-assign coral categories 

```{r}

# Nested if else to create a new variable where genera are sorted into benthic category type.


benthic.cats <- t.cover %>%
  mutate(Category = 
      ifelse(Genera %in% c("Acanthastrea.Bleached", "Acanthastrea.Healthy", "Acanthastra.Healthy","Acanthastrea.Paling"),
             "Acanthastrea",
      ifelse(Genera %in% c("Acropora.Branching.Bleached","Acropora.Branching.Healthy","Acropora.Branching.Paling"), 
             "Acropora_branch",
      ifelse(Genera %in% c("Acropora.NonBranching.Bleached","Acropora.NonBranching.Healthy","Acropora.NonBranching.Paling"), 
             "Acropora_non_branch",
      ifelse(Genera %in% c("Goniopora.Bleached","Goniopora.Healthy","Goniopora.Paling"), 
             "Goniopora",
      ifelse(Genera %in% c("Montipora.Encrusting.Bleached","Montipora.Encrusting.Healthy","Montipora.Encrusting.Paling"),
               "Montipora_encrusting",
      ifelse(Genera %in% c("Montipora.Plating.Bleached","Montipora.Plating.Healthy","Montipora.Plating.Paling"),
             "Montipora_plating", 
      ifelse(Genera %in% c("Other.Branch.Bleached","Other.Branch.Healthy","Other.Branch.Paling"),
             "Pocilloporidae", 
      ifelse(Genera %in% c("Platygyra.Bleached","Platygyra.Healthy","Platygyra.Paling"),
             "Platygyra",   
      ifelse(Genera %in% c("Porites.Bleached","Porites.Healthy","Porites.Paling"),
             "Porites", 
      ifelse(Genera %in% c("Other.Calcified.Invert","Calcifying.algae..Other"),
             "Other Calcifier", 
      ifelse(Genera %in% c("Sand"),
             "Sand",
      ifelse(Genera %in% c("Rubble","Bare_rock"),
             "Rubble and Rock", 
      ifelse(Genera %in% c("Algae..Fleshy.","Kelp"),
             "Fleshy", 
      ifelse(Genera %in% c("Macroalgae"),
             "Macroalgae", 
      ifelse(Genera %in% c("Red.Turf"),
             "Red Turf", 
      ifelse(Genera %in% c("Turf.algae"),
             "Green Turf",
      ifelse(Genera %in% c("Black_turf"),
             "Black Turf",
      ifelse(Genera %in% c("All.other"),
               "Other","NA"
             )))))))))))))))))))

```


```{r}
# Sum groups for each PQ so that we have a total cover per grouping, per time points

str(benthic.cats)

groups_sum <- benthic.cats %>% 
  group_by(TP,Transect_ID,PQ,Category) %>%
  summarise(Total_cover = sum(Cover))

write.csv(groups_sum, "Cover_per_category_per_quadrat.csv")
  
# Search for any NAs

filter(groups_sum, Category == "NA")

# These are here because they are dead coral or the other category (we don't know what that is...probably should remove NAs but leave in for now to see how much cover they take up)
```

Specify the main groups for plotting

```{r}

# Cover per PQ per quadrat

# Split Transect_ID into EB and SB 

groups_sum_split <- groups_sum %>%
  separate(Transect_ID, c("Site", "Transect"), "_") 

cover.groups <- groups_sum_split  %>%
  mutate(Main_groups = 
      ifelse(Category %in% c("Acropora_branch", "Acropora_non_branch", "Acanthastrea","Goniopora","Montipora_encrusting","Montipora_plating","Pocilloporidae","Platygyra","Porites"),
             "Hard_coral",
      ifelse(Category %in% c("Macroalgae"), 
             "Macroalgae",
       ifelse(Category %in% c("Red Turf"), 
             "Red_Turf",
       ifelse(Category %in% c("Green Turf"), 
              "Green_Turf",
        ifelse(Category %in% c("Black Turf"),
               "Black_Turf",
      ifelse(Category %in% c("Rubble and Rock"),
             "Rubble and Rock", 
      ifelse(Category %in% c("Other"),
      "Other","NA"
             ))))))))


# SUM for main groups per PQ 

main_groups_cover_transect <- cover.groups %>% 
  group_by(TP,Site,Transect, PQ,Main_groups) %>% 
  summarise(sum.PQ = sum(Total_cover))

# Average main groups per transect

average_cover_main_groups <- main_groups_cover_transect %>% 
  group_by(TP,Site,Transect,Main_groups)%>% 
  summarise(avg.cover=mean(sum.PQ))


write.csv(average_cover_main_groups, "NFI_transect_average_cover.csv")
            
# AVG cover per TP, Site (SE)
            
average_cover_all <- average_cover_main_groups %>% 
  group_by(TP,Site,Main_groups) %>% 
  summarise(avg.cover.all=mean(avg.cover),
            median_cover = median(avg.cover),
  N = n(),
  SD_cover = sd(avg.cover),
  SE_cover = sd(avg.cover/sqrt(n())))

write.csv(average_cover_all, "NFI_time_site_average_cover.csv")
            

  
```

# Plotting 

Order time points

```{r}

# Order TP 

average_cover_all$TP <- ordered(average_cover_all$TP, levels = c("March_2020","March_2021","April_2022","Sept_2022","Dec_2022","April_2023"))

average_cover_main_groups$TP <- ordered(average_cover_main_groups$TP, levels = c("March_2020","March_2021","April_2022","Sept_2022","Dec_2022","April_2023"))

```

# EB plot 

```{r}

# Subset EB and remove NA

# Box plot 

EB_only_average_cover_all_box <- average_cover_main_groups %>% filter(Site == "EB") %>% subset(Main_groups != "NA")

EB_only_average_cover_all_box $Main_groups <- ordered(EB_only_average_cover_all_box $Main_groups, levels = c("Hard_coral","Macroalgae","Red_Turf","Green_Turf","Black_Turf","Rubble and Rock"))


ggplot(EB_only_average_cover_all_box, aes(x=TP, y=avg.cover)) +
  geom_boxplot() +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

# Remove March 2021 

EB_only_average_cover_all_box_minus_March_2021 <- EB_only_average_cover_all_box %>% filter(TP != "March_2021")

ggplot(EB_only_average_cover_all_box_minus_March_2021, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

# Coral groups

EB_coral <- cover.groups %>%
  filter(Main_groups == "Hard_coral") %>%
  filter(Site == "EB") %>%
  filter(TP != "March_2021")

EB_coral_transect <- EB_coral %>% group_by(TP,Site,Transect,Category) %>%
  summarise(cover = mean(Total_cover))


EB_coral_transect$Category <- ordered(EB_coral_transect$Category, levels = c("Acropora_branch","Acropora_non_branch","Montipora_encrusting","Montipora_plating","Pocilloporidae","Acanthastrea","Goniopora","Platygyra","Porites"))

EB_coral_transect$TP <- ordered(EB_coral_transect$TP, levels = c("March_2020","April_2022","Sept_2022","Dec_2022","April_2023"))

ggplot(EB_coral_transect, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

```

# SB plot 


```{r}

# Box plot 

SB_only_average_cover_all_box <- average_cover_main_groups %>% filter(Site == "SB") %>% subset(Main_groups != "NA")

SB_only_average_cover_all_box $Main_groups <- ordered(SB_only_average_cover_all_box $Main_groups, levels = c("Hard_coral","Macroalgae","Red_Turf","Green_Turf","Black_Turf","Rubble and Rock"))

average_cover_main_groups 


ggplot(SB_only_average_cover_all_box, aes(x=TP, y=avg.cover)) +
  geom_boxplot() +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

SB_only_average_cover_all_box_minus_March_2021 <- SB_only_average_cover_all_box %>% filter(TP != "March_2021")

ggplot(SB_only_average_cover_all_box_minus_March_2021, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

# Coral groups

SB_coral <- cover.groups %>%
  filter(Main_groups == "Hard_coral") %>%
  filter(Site == "SB") %>%
  filter(TP != "March_2021")

SB_coral_transect <- SB_coral %>% group_by(TP,Site,Transect,Category) %>%
  summarise(cover = mean(Total_cover))


SB_coral_transect$Category <- ordered(SB_coral_transect$Category, levels = c("Acropora_branch","Acropora_non_branch","Montipora_encrusting","Montipora_plating","Pocilloporidae","Acanthastrea","Goniopora","Platygyra","Porites"))

SB_coral_transect$TP <- ordered(SB_coral_transect$TP, levels = c("March_2020","April_2022","Sept_2022","Dec_2022","April_2023"))

ggplot(SB_coral_transect, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

```

# Plotting of cover groups in SB (SB East, SB Middle, SB West (stairs))

```{r}


SB_transects_maingroups <- main_groups_cover_transect %>% filter(Site == "SB") %>% 
  group_by(TP,Site,Transect,Main_groups)%>% 
  summarise(avg.cover=mean(sum.PQ))

SB_transects_maingroups$TP <- ordered(SB_transects_maingroups$TP, levels = c("March_2020","March_2021","April_2022","Sept_2022","Dec_2022","April_2023"))

SB_transects_maingroups$Main_groups <- ordered(SB_transects_maingroups$Main_groups, levels = c("Hard_coral","Macroalgae","Red_Turf","Green_Turf","Black_Turf","Rubble and Rock"))

SB_main_groups_subsetted <- SB_transects_maingroups %>% 
  filter(TP != "March_2021") %>%
  mutate(Transect_group = 
      ifelse(Transect %in% c("9", "8","10","11","6","12","5"),
             "SB_west",
      ifelse(Transect %in% c("4","3","24","26","25","27","14","13","15","16","17"),
             "SB_middle",
       ifelse(Transect %in% c("1","23","22","21","18","19","20"), 
             "SB_East","NA"
             ))))

SB_west_main <- SB_main_groups_subsetted %>% filter(Transect_group == "SB_west") %>% filter(Main_groups != "NA") 

ggplot(SB_west_main, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

SB_middle_main <- SB_main_groups_subsetted %>% filter(Transect_group == "SB_middle") %>% filter(Main_groups != "NA") 

ggplot(SB_middle_main, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

SB_East_main <- SB_main_groups_subsetted %>% filter(Transect_group == "SB_East") %>% filter(Main_groups != "NA") 

ggplot(SB_East_main, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)



```

# Plotting of coral cover groups in SB (SB East, SB Middle, SB West (stairs))

```{r}

# Coral groups

SB_coral <- cover.groups %>%
  filter(Main_groups == "Hard_coral") %>%
  filter(Site == "SB") %>%
  filter(TP != "March_2021")

SB_coral_transect <- SB_coral %>% group_by(TP,Site,Transect,Category) %>%
  summarise(cover = mean(Total_cover))

SB_coral_transect$Category <- ordered(SB_coral_transect$Category, levels = c("Acropora_branch","Acropora_non_branch","Montipora_encrusting","Montipora_plating","Pocilloporidae","Acanthastrea","Goniopora","Platygyra","Porites"))

SB_coral_transect$TP <- ordered(SB_coral_transect$TP, levels = c("March_2020","April_2022","Sept_2022","Dec_2022","April_2023"))

SB_coral <- SB_coral_transect %>%
  mutate(Transect_group = 
      ifelse(Transect %in% c("9", "8","10","11","6","12","5"),
             "SB_west",
      ifelse(Transect %in% c("4","3","24","26","25","27","14","13","15","16","17"),
             "SB_middle",
       ifelse(Transect %in% c("1","23","22","21","18","19","20"), 
             "SB_East","NA"
             ))))

SB_west_coral <- SB_coral %>% filter(Transect_group == "SB_west")

ggplot(SB_west_coral, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

SB_middle_coral <- SB_coral %>% filter(Transect_group == "SB_middle")

ggplot(SB_middle_coral, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

SB_East_coral <- SB_coral %>% filter(Transect_group == "SB_East")

ggplot(SB_East_coral, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)



```