---
title: "NFI LTMP Benthic Plots"
author: "Shannon Eckhardt"
date: "2023-10-12"
output: html_document
---

# LOAD LIBRARIES & DATA
```{r}
rm(list = ls())
library(tidyverse)
library(readxl)


master_data <- read_excel("/Users/Shannon/GitHub/NFI_LTMP/01_Data/Benthic_cover_March20-March23.xlsx")
# no september 2022 in this data

```

# DATA CLEANING

```{r}
# Pick out site, transect_ID and PQ
master_data <- master_data %>%
  mutate(Site = str_extract(Image_ID, "[A-Z]{2}"),
         Transect_ID = str_extract(Image_ID, "[A-Z]{2}\\d+(?=_)"),
  # extract the number before the end of string
         PQ = str_extract(Image_ID, "\\d+$")) %>% 
  dplyr::select(Image_ID, TP, Site, Transect_ID, PQ, everything()) # reorder columns

#unique(master_data$Site)
```

# CATEGORIES

Transpose data
```{r}
t.cover <- master_data %>% pivot_longer(c(ABH:`T`), names_to = "Genera", values_to = "Cover")
```
Grouping into broader categories
```{r}
benthic.cats <- t.cover %>%
  mutate(Category = 
      ifelse(Genera %in% c("ACH"),
             "Acanthastrea",
      ifelse(Genera %in% c("ABH"), 
             "Branching Acropora",
      ifelse(Genera %in% c("ANH"), 
             "Plating Acropora",
      ifelse(Genera %in% c("GOH"), 
             "Goniopora",
      ifelse(Genera %in% c("MEH"),
             "Encrusting Montipora",
      ifelse(Genera %in% c("MPH"),
             "Plating Montipora", 
      ifelse(Genera %in% c("OBH"),
             "Pocilloporidae", 
      ifelse(Genera %in% c("PLH"),
             "Platygyra",   
      ifelse(Genera %in% c("POH"),
             "Porites", 
      ifelse(Genera %in% c("OG"),
             "Other hard coral", 
      ifelse(Genera %in% c("S"),
             "Sand",
      ifelse(Genera %in% c("R","BARK"),
             "Rubble and Rock", 
      ifelse(Genera %in% c("Flesh", "Fol", "Dictyota", "Caulerpa", "Leath", "ACA", "CCA"),
             "Macroalgae", 
      ifelse(Genera %in% c("RT"),
             "Red Turf", 
      ifelse(Genera %in% c("T"),
             "Green Turf",
      ifelse(Genera %in% c("OtT"),
             "Other Turf",
      ifelse(Genera %in% c("BT"),
             "Black Turf",
      ifelse(Genera %in% c("ASCD", "anemo", "ESCB", "soft", "Urchin"),
             "Other benthic invertebrates",
      ifelse(Genera %in% c("NA"),
               "Other","NA"
             ))))))))))))))))))))
```
SUM counts/cover per quadrat
```{r}
groups_sum <- benthic.cats %>% 
  group_by(TP,Site,Transect_ID,PQ,Category) %>%
  summarise(Total_cover = sum(Cover))

write.csv(groups_sum, file = "/Users/Shannon/GitHub/NFI_LTMP/01_Data/Cover_per_category_per_quadrat.csv")

# search for NAs
filter(groups_sum, Category == "NA")

# no NAs because dead coral is now in the category R = Rubble
```

Specify the main groups for plotting
- Hard coral as one category and not separate genera
```{r}
cover.groups <- groups_sum  %>%
  mutate(Main_groups = 
      ifelse(Category %in% c("Branching Acropora", "Plating Acropora", "Acanthastrea","Goniopora","Encrusting Montipora","Plating Montipora","Pocilloporidae","Platygyra","Porites", "Other hard coral"),
             "Hard coral",
      ifelse(Category %in% c("Macroalgae"), 
             "Macroalgae",
      ifelse(Category %in% c("Other Turf"), 
             "Other Turf",
      ifelse(Category %in% c("Red Turf"), 
             "Red Turf",
      ifelse(Category %in% c("Green Turf"), 
              "Green Turf",
      ifelse(Category %in% c("Black Turf"),
               "Black Turf",
      ifelse(Category %in% c("Rubble and Rock"),
             "Rubble and Rock", 
      ifelse(Category %in% c("Sand"),
             "Sand",
      ifelse(Category %in% c("Other benthic invertebrates"),
             "Other benthic invertebrates",
      ifelse(Category %in% c("Other"),
      "Other","NA"
             )))))))))))

# SUM for main groups per quadrat
main_groups_cover_transect <- cover.groups %>% 
  group_by(TP,Site,Transect_ID, PQ, Main_groups) %>% 
  summarise(sum.PQ = sum(Total_cover))

# AVG main groups per transect
average_cover_main_groups <- main_groups_cover_transect %>% 
  group_by(TP,Site,Transect_ID,Main_groups)%>% 
  summarise(avg.cover=mean(sum.PQ))

write.csv(average_cover_main_groups, file = "/Users/Shannon/GitHub/NFI_LTMP/01_Data/NFI_transect_average_cover.csv")

# AVG cover per TP, Site (SE)
            
average_cover_all <- average_cover_main_groups %>% 
  group_by(TP,Site,Main_groups) %>% 
  summarise(avg.cover.all=mean(avg.cover),
            median_cover = median(avg.cover),
  N = n(),
  SD_cover = sd(avg.cover),
  SE_cover = sd(avg.cover/sqrt(n())))

write.csv(average_cover_all, "/Users/Shannon/GitHub/NFI_LTMP/01_Data/NFI_time_site_average_cover.csv")

# AVG cover per TP, Site (SE) for hard coral cover -----------------------------------------------------------
coral.groups <- cover.groups %>% subset(Main_groups == "Hard coral")

# SUM for coral groups per quadrat
coral_cover_transect <- coral.groups %>% 
  group_by(TP,Site,Transect_ID, PQ, Category) %>% 
  summarise(sum.PQ = sum(Total_cover))

# AVG coral groups per transect
average_cover_coral_groups <- coral_cover_transect %>% 
  group_by(TP,Site,Transect_ID,Category)%>% 
  summarise(avg.cover=mean(sum.PQ))

# AVG cover per TP, Site (SE)
average_cover_all_coral <- average_cover_coral_groups %>% 
  group_by(TP,Site,Category) %>% 
  summarise(avg.cover.all=mean(avg.cover),
            median_cover = median(avg.cover),
  N = n(),
  SD_cover = sd(avg.cover),
  SE_cover = sd(avg.cover/sqrt(n())))

write.csv(average_cover_all_coral, "/Users/Shannon/GitHub/NFI_LTMP/01_Data/NFI_time_site_average_coral_cover.csv")

```


# PLOTTING

Order time points
```{r}

# Order TPs

average_cover_all$TP <- ordered(average_cover_all$TP, levels = c("March_2020","Dec_2020", "March_2021","April_2022","Dec_2022","March_2023"))

average_cover_main_groups$TP <- ordered(average_cover_main_groups$TP, levels = c("March_2020","Dec_2020","March_2021","April_2022","Dec_2022","March_2023"))

```

## EB PLOT

```{r}

# Subset EB

# Box plot 

EB_only_average_cover_all_box <- average_cover_main_groups %>% filter(Site == "EB") %>% subset(Main_groups != c("Other", "Sand")) # remove sand and other category

EB_only_average_cover_all_box$Main_groups <- ordered(EB_only_average_cover_all_box$Main_groups, levels = c("Hard coral","Macroalgae","Red Turf","Green Turf","Black Turf", "Other Turf", "Other benthic invertebrates","Rubble and Rock"))


EB_cats <- ggplot(EB_only_average_cover_all_box, aes(x=TP, y=avg.cover)) +
  geom_boxplot() +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/EB_major_cats_cover.jpg", plot = EB_cats, width = 14, height = 4, dpi = 300)

# CORAL GENERA -----------------------------------------------------------------------------------------------------

EB_coral <- cover.groups %>%
  filter(Main_groups == "Hard coral") %>%
  filter(Site == "EB")

EB_coral_transect <- EB_coral %>% group_by(TP,Site,Transect_ID,Category) %>%
  summarise(cover = mean(Total_cover))


EB_coral_transect$Category <- ordered(EB_coral_transect$Category, levels = c("Branching Acropora","Plating Acropora","Encrusting Montipora","Plating Montipora","Pocilloporidae","Acanthastrea","Goniopora","Platygyra","Porites", "Other hard coral"))

EB_coral_transect$TP <- ordered(EB_coral_transect$TP, levels = c("March_2020", "Dec_2020", "March_2021", "April_2022","Dec_2022","March_2023"))

EB_coral <- ggplot(EB_coral_transect, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/EB_coral_cover.jpg", plot = EB_coral, width = 14, height = 4, dpi = 300)

```

# SB PLOT
```{r}
# Box plot 

SB_only_average_cover_all_box <- average_cover_main_groups %>% filter(Site == "SB") %>% subset(Main_groups != c("Other", "Sand")) # remove sand and other category

SB_only_average_cover_all_box$Main_groups <- ordered(SB_only_average_cover_all_box $Main_groups, levels = c("Hard coral","Macroalgae","Red Turf","Green Turf","Black Turf", "Other Turf", "Other benthic invertebrates","Rubble and Rock"))

SB_cats <- ggplot(SB_only_average_cover_all_box, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size=0.5) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/SB_major_cats_cover.jpg", plot = SB_cats, width = 14, height = 4, dpi = 300)


# CORAL GENERA -----------------------------------------------------------------------------------------------------------

SB_coral <- cover.groups %>%
  filter(Main_groups == "Hard coral") %>%
  filter(Site == "SB")

SB_coral_transect <- SB_coral %>% group_by(TP,Site,Transect_ID,Category) %>%
  summarise(cover = mean(Total_cover))


SB_coral_transect$Category <- ordered(SB_coral_transect$Category, levels = c("Branching Acropora","Plating Acropora","Encrusting Montipora","Plating Montipora","Pocilloporidae","Acanthastrea","Goniopora","Platygyra","Porites", "Other hard coral"))

SB_coral_transect$TP <- ordered(SB_coral_transect$TP, levels = c("March_2020","Dec_2020", "March_2021", "April_2022","Dec_2022","March_2023"))

SB_coral <- ggplot(SB_coral_transect, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/SB_coral_cover.jpg", plot = SB_coral, width = 14, height = 4, dpi = 300)

```
# PLOTTING SB EAST, MIDDLE, WEST (STAIRS)

Transects per time point:

March 2020: 7 SB ----- EXCLUDE, as transects are different and there is no transect for SB middle
Dec 2020: 20 SB ------ CAN'T FIND MAP - include for now
March 2021: 25 SB
April 2022: 25 SB (missing 2 & 7)
Dec 2022: 24 SB (missing 2, 7 & 24)
March 2023: 25 SB


```{r}
# exclude March 2020
SB_avg_cover_main_groups <- average_cover_main_groups %>% 
  filter(Site == "SB") %>% 
  filter(TP != "March_2020") %>% 
  filter(Main_groups != "Sand") %>%
  filter(Main_groups != "Other")

# unique(SB_avg_cover_main_groups$TP)

SB_avg_cover_main_groups$TP <- ordered(SB_avg_cover_main_groups$TP, levels = c("Dec_2020","March_2021","April_2022","Dec_2022","March_2023"))

SB_avg_cover_main_groups$Main_groups <- ordered(SB_avg_cover_main_groups$Main_groups, levels = c("Hard coral","Macroalgae","Red Turf","Green Turf","Black Turf","Other Turf","Other benthic invertebrates","Rubble and Rock"))

# Remove leading zeros from Transect_ID
SB_avg_cover_main_groups$Transect_ID <- sub("^(SB|CB|EB)(0+)([1-9])$", "\\1\\3", SB_avg_cover_main_groups$Transect_ID)

# unique(SB_avg_cover_main_groups$Transect_ID)

# SB2 and SB7 for Dec 2020 not included
SB_main_groups_subsetted <- SB_avg_cover_main_groups %>% 
  mutate(Transect_group = 
      ifelse(Transect_ID %in% c("SB9", "SB8","SB10","SB11","SB6","SB12","SB5"),
             "SB_west",
      ifelse(Transect_ID %in% c("SB4","SB3","SB24","SB26","SB25","SB27","SB14","SB13","SB15","SB16","SB17"),
             "SB_middle",
       ifelse(Transect_ID %in% c("SB1","SB23","SB22","SB21","SB18","SB19","SB20"), 
             "SB_East","NA"
             ))))

# SB WEST --------------------------------------------
SB_west_main <- SB_main_groups_subsetted %>% filter(Transect_group == "SB_west") %>% filter(Main_groups != "NA") 

SB_west <- ggplot(SB_west_main, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/SB_west_major_cats_cover.jpg", plot = SB_west, width = 14, height = 4, dpi = 300)

# SB MIDDLE ------------------------------------------
SB_middle_main <- SB_main_groups_subsetted %>% filter(Transect_group == "SB_middle") %>% filter(Main_groups != "NA") 

SB_middle <- ggplot(SB_middle_main, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/SB_middle_major_cats_cover.jpg", plot = SB_middle, width = 14, height = 4, dpi = 300)

# SB EAST --------------------------------------------
SB_East_main <- SB_main_groups_subsetted %>% filter(Transect_group == "SB_East") %>% filter(Main_groups != "NA") 

SB_east <- ggplot(SB_East_main, aes(x=TP, y=avg.cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,100)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Main_groups)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/SB_east_major_cats_cover.jpg", plot = SB_east, width = 14, height = 4, dpi = 300)




# make file to export
# remove NAs
SB_main_groups_subsetted_noNA <- subset(SB_main_groups_subsetted, Transect_group != "NA")
# because December 2020 and March 2021 have SB2 and SB7

# AVG cover per TP, Site (SE)
average_cover_SB_main_groups_subsetted <- SB_main_groups_subsetted_noNA %>% 
  group_by(TP,Transect_group,Main_groups) %>% 
  summarise(avg.cover.all=mean(avg.cover),
            median_cover = median(avg.cover),
  N = n(),
  SD_cover = sd(avg.cover),
  SE_cover = sd(avg.cover/sqrt(n())))

write.csv(average_cover_SB_main_groups_subsetted, "/Users/Shannon/GitHub/NFI_LTMP/01_Data/NFI_time_site_average_main_groups_cover_SB_sites.csv")

```


# Plotting of coral cover groups in SB (SB East, SB Middle, SB West (stairs))
```{r}
SB_coral <- cover.groups %>%
  filter(Main_groups == "Hard coral") %>%
  filter(Site == "SB") %>%
  filter(TP != "March_2020")

SB_coral_transect <- SB_coral %>% group_by(TP,Site,Transect_ID,Category) %>%
  summarise(cover = mean(Total_cover))

SB_coral_transect$Category <- ordered(SB_coral_transect$Category, levels = c("Branching Acropora","Plating Acropora","Encrusting Montipora","Plating Montipora","Pocilloporidae","Acanthastrea","Goniopora","Platygyra","Porites", "Other hard coral"))

# exclude March 2020
SB_coral_transect$TP <- ordered(SB_coral_transect$TP, levels = c("Dec_2020", "March_2021", "April_2022","Dec_2022","March_2023"))

# Remove leading zeros from Transect_ID
SB_coral_transect$Transect_ID <- sub("^(SB|CB|EB)(0+)([1-9])$", "\\1\\3", SB_coral_transect$Transect_ID)

# SB2 and SB7 for Dec 2020 not included
SB_coral <- SB_coral_transect %>% 
  mutate(Transect_group = 
      ifelse(Transect_ID %in% c("SB9", "SB8","SB10","SB11","SB6","SB12","SB5"),
             "SB_west",
      ifelse(Transect_ID %in% c("SB4","SB3","SB24","SB26","SB25","SB27","SB14","SB13","SB15","SB16","SB17"),
             "SB_middle",
       ifelse(Transect_ID %in% c("SB1","SB23","SB22","SB21","SB18","SB19","SB20"), 
             "SB_East","NA"
             ))))

# SB WEST CORAL --------------------------------------------
SB_west_coral <- SB_coral %>% filter(Transect_group == "SB_west")

SB_west_cor <- ggplot(SB_west_coral, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/SB_west_coral_cover.jpg", plot = SB_west_cor, width = 14, height = 4, dpi = 300)

# SB MIDDLE CORAL --------------------------------------------
SB_middle_coral <- SB_coral %>% filter(Transect_group == "SB_middle")

SB_middle_cor <- ggplot(SB_middle_coral, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/SB_middle_coral_cover.jpg", plot = SB_middle_cor, width = 14, height = 4, dpi = 300)

# SB EAST CORAL --------------------------------------------
SB_East_coral <- SB_coral %>% filter(Transect_group == "SB_East")

SB_east_cor <- ggplot(SB_East_coral, aes(x=TP, y=cover)) +
  geom_boxplot(outlier.size = 0.1) +
  theme_bw() +
  xlab("") +
  ylab("Cover (%)") +
  scale_y_continuous(expand = c(0, 0), limits = c(0,60)) +
  theme(axis.text.x = element_text(hjust=0.95,vjust=0.2,angle = 90)) +
    facet_grid(~Category)

ggsave(filename = "/Users/Shannon/GitHub/NFI_LTMP/03_Plots/SB_east_coral_cover.jpg", plot = SB_east_cor, width = 14, height = 4, dpi = 300)



# make file to export
# remove NAs
SB_coral_noNA <- subset(SB_coral, Transect_group != "NA")
# because December 2020 and March 2021 have SB2 and SB7

# AVG cover per TP, Site (SE)
average_cover_SB_coral_subsetted <- SB_coral_noNA %>% 
  group_by(TP,Transect_group,Category) %>% 
  summarise(avg.cover.all=mean(cover),
            median_cover = median(cover),
  N = n(),
  SD_cover = sd(cover),
  SE_cover = sd(cover/sqrt(n())))

write.csv(average_cover_SB_coral_subsetted, "/Users/Shannon/GitHub/NFI_LTMP/01_Data/NFI_time_site_average_coral_cover_SB_sites.csv")


```

